---
title: "Google Data Analytics Capstone"
author: "Chen"
date: '2022-06-16'
output:
  pdf_document: default
  html_document: default
---

```{r}

```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load the necessary packages

```{r message=FALSE, warning=FALSE, , message=FALSE, paged.print=FALSE}
library(tidyverse)
library(skimr)
library(lubridate)
library(hms)
library(janitor)
```

## Step 1: Import Data

Load csv files here

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
trip_202106 <- read_csv("202106-divvy-tripdata.csv")
trip_202107 <- read_csv("202107-divvy-tripdata.csv")
trip_202108 <- read_csv("202108-divvy-tripdata.csv")
trip_202109 <- read_csv("202109-divvy-tripdata.csv")
trip_202110 <- read_csv("202110-divvy-tripdata.csv")
trip_202111 <- read_csv("202111-divvy-tripdata.csv")
trip_202112 <- read_csv("202112-divvy-tripdata.csv")
trip_202201 <- read_csv("202201-divvy-tripdata.csv")
trip_202202 <- read_csv("202202-divvy-tripdata.csv")
trip_202203 <- read_csv("202203-divvy-tripdata.csv")
trip_202204 <- read_csv("202204-divvy-tripdata.csv")
trip_202205 <- read_csv("202205-divvy-tripdata.csv")
```

Check whether the data frames created will successfully bind together by rows

While the column names don't have to be in the exact same order, the data frames need to match perfectly with one another before merging them into a single one

If not, rename columns to make them consistent with each other

If necessary, change data types of particular columns so that they can stack correctly (when using bind_rows function)

```{r}
compare_df_cols_same (trip_202106, trip_202107, trip_202108, trip_202109, 
                      trip_202110, trip_202111, trip_202112, trip_202201, 
                      trip_202202, trip_202203, trip_202204, trip_202205,
                      bind_method = 'rbind')
```

Inspect the data frames and look for inconsistencies

```{r}
str(trip_202106)
str(trip_202107)
str(trip_202108)
str(trip_202109)
str(trip_202110)
str(trip_202111)
str(trip_202112)
str(trip_202201)
str(trip_202202)
str(trip_202203)
str(trip_202204)
str(trip_202205)
```

Merge the data frames into one

```{r}
merged_trip <- rbind(trip_202106, trip_202107, trip_202108, trip_202109, 
                         trip_202110, trip_202111, trip_202112, trip_202201, 
                         trip_202202, trip_202203, trip_202204, trip_202205)
```

## Step 2: Clean and Organize Data

Inspect the data frame that has been created

```{r}
colnames(merged_trip) # returns the column names of the data frame
nrow(merged_trip) # returns the number of records in the data frame
head(merged_trip) # returns the first 6 rows of the data frame
str(merged_trip) # returns the internal structure of the data frame
skim_without_charts(merged_trip) # returns summary statistics of the data frame
```

In the data frame there exist entries where rider_length is negative

```{r}
merged_trip <- merged_trip[!(merged_trip$ride_length < 0),]
```

Examine whether there exist any missing dates in the observed period of the data

```{r}
daterange <- seq(min(as.Date(merged_trip$started_at)), 
                 max(as.Date(merged_trip$started_at)), by = "day")
daterange[!daterange %in% as.Date(merged_trip$started_at)]
```

Check whether there are problems that need to be fixed

## Step 3: Transform Data

Create columns that list the date, month, day, and year of each ride

```{r}
merged_trip$date <- as.Date(format(merged_trip$started_at, "%Y%m%d"), "%Y%m%d") 
merged_trip$month <- format(merged_trip$started_at, "%m")
merged_trip$day <- format(merged_trip$started_at, "%d")
merged_trip$year <- format(merged_trip$started_at, "%Y")

```

Create a column named "ride_length." Calculate the length of each ride by subtracting the column "started_at" from the column "ended_at" and format as HH:MM:SS

```{r}
merged_trip <- merged_trip %>% 
  mutate(ride_length = as.numeric(difftime(ended_at, started_at, units = "secs")))

```

Create a column called "day_of_week." Calculate the day of the week that each ride started

```{r}
Sys.setlocale("LC_TIME", "English")
merged_trip$day_of_week <- weekdays(merged_trip$started_at)
```

Create a column called "hour". Calculate the hour of the day that each ride started

```{r}
merged_trip <- merged_trip %>% 
  mutate(hour = hour(started_at))
```

## Export the data to csv file for further analysis

```{r}
write.csv(merged_trip, "merged_tripdata.csv")
```
